---
title: "Hydrology "
output: html_notebook
---

#### The Hydrology Session A

This session contains some elementary Exploratory data analysis (EDA) procedures. 
These procedures contain basic plotting and representative statistics computing.

The main recent data analysis development in R evolves around the `tidyverse` 
package. The package itself is not a single package with distinct functionality, 
rather is it a collection of packages with the same shared philosophy.

```{r setup}
# check if the package needs to be installed, then load the package
for(i in c("tidyverse", "data.table", "lubridate")) {
  if(!(i %in% installed.packages()[, 1])) install.packages(i)
  library(i, character.only = TRUE)
}
```

Now specify url and download the file from that location.

```{r file download}
url <- "https://hydrology.nws.noaa.gov/pub/gcip/mopex/US_Data/Us_438_Daily/01200000.dly" # this is the unique web location of the file
dt <- read.fwf(file = url,                                                               # we are reading the fixed width column format
               widths = c(8, rep(10, times = 5)),                                        # so we need to specify number and width of the columns
               col.names = c("Date", "Prec", "PET", "Discharge", "Tmax", "Tmin")) |>     # it is convenient to name them while loading
  as.data.table()
```

We downloaded and loaded the data set directly using website URL. Since it is a 
fixed width format, we need to use specific function in order to properly load 
the format. Now we need to adjust the data a bit. R does now how to treat Date 
format. If specified we gain the ability to conveniently aggregate and process 
and it is also a good practice to lock the values in specified assumptions 
as much as possible.

```{r date reshape}
dt <- dt |> 
  mutate(Date = gsub(x = Date, pattern = " ", replacement = "0")) |>            # first find and replace all empty spaces in the raw character format
  mutate(Date = as.Date(Date, format = "%Y%m%d"))                               # change the character to Date by specifying Date internal formatting
```

Now we can work on some exploratory data analysis processes (visualization, statistics, tests). 
Within the `tidyverse` package space a pipe operator ` |> ` plays vital role. 
It bear the logic of sending results of previous function call as data input to next step.
At first it is necessary to check the consistency and quality of the data and act
accordingly.

```{r checks}
dt |> 
  sapply(FUN = range)                                                           # simplified apply of range function on dt variables  
```

Two things can be seen. The Date has seemingly wrong values, but that is due to 
internal representation of the datetime format, so it is actually not wrong. But 
the secondly the minimum of Discharge is -99 and that tackles a clear physical 
boundary, so we can assume that the value represents missing values or flawed
measurements and we treat them.

```{r}
dt <- dt |> 
  mutate(Discharge = na_if(Discharge, "-99"))

dt |> ggplot(aes(x = Date, y = Discharge)) +
  geom_line() +
  theme_minimal() +
  ylab("Discharge [mm]")
```


### Hydrological year

In the Czechia it is defined form the first of November to the last of October.

```{r HYr definition}
dt <- dt |> 
  mutate(HYr = ifelse(month(Date) > 10, 
                      year(Date) + 1, 
                      year(Date)))
```




```{r}
dt |> 
  group_by(lubridate::year(Date)) |>                                            # for each year
  summarize(n = n()) |>                                                         # let's count entries
  distinct(n)                                                                   # there is a full time series, just observations are missing

dt |>                                                                     # reduce dt to only those years with full records of Disch
  filter(!year(Date) %in% {
    unique(dt[is.na(Discharge), HYr])  
})
```


```{r}
dt |>                                                                           # pipe the dataset further to grouping function 
  group_by(Mon = lubridate::month(Date)) |>                                     # begin with grouping to months since our aim is monthly stats
  summarize(Prec = mean(Prec)) |>                                               # summarize function allows computations on selected variables
  ggplot() +
  geom_col(aes(x = as.factor(Mon), 
               y = Prec, 
               group = 1), 
           width = 0.2, 
           color = "darkblue") +
  scale_x_discrete(labels = 1:12, breaks = 1:12) +                              # months are integers - so we specify x-axis format 
  theme_minimal() +                                                             # adopt visually simple theme
  xlab("Month") +                                                               # changes the label of x-axis 
  ylab("Mean precipitation total [mm]")                                         # changes the label of y-axis 
  
```


```{r}
dt |> 
  group_by(Mon = lubridate::month(Date), 
           Year = lubridate::year(Date)) |> 
  summarize(Tmin = mean(Tmin), 
            Tmax = mean(Tmax)) |> 
  pivot_longer(-c(Mon, Year), names_to = "Variable") |> 
  ggplot(aes(x = as.factor(Mon), 
             y = value, 
             color = Variable)) +
  theme_minimal() +
  geom_boxplot() +
  xlab("Month") +
  ylab("Temperature [\u00B0C]") +
  scale_color_manual(values = c("orange", "dodgerblue3"))
```

$$
\varphi = \dfrac{\text{total runoff}}{\text{total precipitation}}\quad\text{(per year)}
$$

```{r runoff coefficient}
# since we are missing some days with measurements, we would like to include only complete years into our analysis
dt |> 
  filter(!HYr %in% {
    unique(dt[is.na(Discharge), HYr])  
}) |> 
  group_by(HYr) |> 
  mutate(Yr_Prec = sum(Prec), Yr_Disch = sum(Discharge), phi = Yr_Disch/Yr_Prec) |> 
  # ungroup() |> 
  distinct(HYr, phi) |> 
  ggplot(aes(x = HYr, y = phi)) +
  geom_col() +
  scale_x_date("1 year")
```


